Approximation as Differentiation

### 2025-04-29 appendix/notes/auto-diff.tex
Backward derivative.
Basic AD setup.

Write $\tangents_x(\RR^n)$ for the tangent space at a point $x \in \RR^n$. Then the \emph{forward derivative}
${\pushf{f}}_x$ (tangent map or pushforward) of a function $f: \RR^m \to \RR^n$ at $x$ is a linear map
$\tangents_x(\RR^m) \linearto \tangents_{f(x)}(\RR^n)$. Because $\RR^n$ is a Euclidean manifold,
$\tangents_x(\RR^n)$ is naturally isomorphic to $\RR^n$.

>>>>

Write $\tangents_x(\RR^n)$ for the tangent space at a point $x \in \RR^n$. Then the \emph{forward derivative}
(tangent map or pushforward) $\pushf{f}_x$ of a differentiable function $f: \RR^m \to \RR^n$ at $x \in
\RR^m$ is a linear map $\tangents_x(\RR^m) \linearto \tangents_{f(x)}(\RR^n)$. Because $\RR^n$ is a Euclidean
manifold, $\tangents_x(\RR^n)$ is naturally isomorphic to $\RR^n$, so actually $\pushf{f}_x: \RR^m \linearto
\RR^n$. The \emph{backward derivative} (cotangent map or pullback) $\pullf{f}_x$  is a linear map $\RR^n
\linearto \RR^m$.

\subsubsection{Chain rule}

Suppose $f: \RR^m \to \RR^n$ and $g: \RR^n \to \RR^k$. For any $x \in \RR^m$ we can define the following
composite derivatives of $g \comp f: \RR^m \to \RR^k$:

\begin{itemize}
\item $\pushf{(g \comp f)}_x = \pushf{g}_{f(x)} \comp \pushf{f}_x$
\item $\pullf{(g \comp f)}_x = \pullf{f}_{x} \comp \pullf{g}_{f(x)}$
\end{itemize}

\begin{definition}[Forward mode automatic differentiation functor]
Define the functor $\tangents_*$ which sends every vector space $\RR^n$ to itself and every smooth map $f:
\RR^m \to \RR^n$ to the function $\tangents_*(f) = x \mapsto (f(x), \pushf{f}_x): \RR^m \to \RR^n \times (\RR^m
\linearto \RR^n)$, associating to every point its image in $f$ and the forward derivative of $f$ at that
point.
\end{definition}

For any map $f: \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$, define $f_1 = \pi_1 \comp f$ and $f_2$ for
$\pi_2 \comp f$. Regardless of whether the linear maps $\RR^m \linearto \RR^n$ happen to be derivatives, we
can define composition on such maps as:
\begin{align*}
(g \comp f)_1(x) &= f_1(g_1(x)) \\
(g \comp f)_2(x) &= g_2(f_1(x)) \comp f_2(x)
\end{align*}
\noindent and verify that $\tangents$ is indeed a functor. If $f_2(x)$ happens to be the derivative of $f_1$
at $x$ for any $x \in \RR^m$ and similarly for $g$ then the derivatives compose according to the forward chain
rule and $(g \comp f)_2(x)$ is the derivative of $(g \comp f)_1$ at $x$.

\begin{definition}[Reverse mode automatic differentiation functor]
Define the functor $\tangents^*$ which also sends every vector space $\RR^n$ to itself but which sends every
smooth map $f: \RR^m \to \RR^n$ to the function $\tangents^*(f) = x \mapsto (f(x), \pullf{f}_x): \RR^m \to
\RR^n \times (\RR^n \to \RR^m)$.
\end{definition}

\noindent and again we can define a notion of composition for such maps that respects the (backward) chain
rule.
